Are translation equivalents special? Evidence from simulations and empirical data from bilingual infants

### 2021-12-13 paper/paper_TE_bilingual_vocabulary_model.Rmd
Updated paper

* Prediction 1a: Children with more balanced vocabularies (BALANCE) will produce more translation equivalents (TE).
* Prediction 1b: Children who produce more total words (WORD) or more dominant-language words (DOM) will produce more translation equivalents (TE). 
* Prediction 1c: Children who produce more non-dominant language words (NONDOM) will produce more translation equivalents (TE); but unlike for WORD and DOM this does not interact with BALANCE; instead, non-dominant vocabulary size will be an almost perfect predictor of translation equivalent knowledge (see panel 1B of Figure 1).

@@

## visualizing Simulation 1
simulation1 <- data_simulation1 %>%
  # convert to long data
  pivot_longer(c(WORD, DOM, NONDOM), names_to = "vocab_type", values_to = "number") %>%
  mutate(vocab_type = factor(vocab_type, levels=c("DOM", "NONDOM", "WORD"))) %>%
  mutate(vocab_type = recode(vocab_type, 
                             DOM = "Panel 1A: \n Dominant vocabulary \n (DOM)", 
                             NONDOM = "Panel 1B: \n Non-dominant vocabulary \n (NONDOM)",
                             WORD = "Panel 1C: \n Total vocabulary \n (WORD)")) %>%
  ggplot(aes(x = number, y = TE, color = bin_BALANCE, linetype = bin_BALANCE)) +
  stat_smooth(method = lm, se = F) +
  scale_color_manual(values=c("#bdbdbd", "#737373", "#525252", "#252525", "#000000")) + 
  scale_linetype_manual(values=c("longdash", "dotdash", "dashed", "twodash", "solid")) +
  theme_minimal() + 
  facet_grid(. ~ vocab_type) +
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom",
        plot.title = element_text(face = 'bold', size = 12, hjust = 0, vjust = 2),
        plot.title.position = "plot",
        text = element_text(size=12),
        panel.spacing.x = unit(2, "lines"),
        panel.background = element_rect(fill = NA, color = "black")) +
  ylim(0, 600) +
  xlim(0, 1000) +
  labs(x = "Number of words", 
       y = "Number of translation equivalents (TE)", 
       color = "Vocabulary balance (BALANCE)",
       linetype = "Vocabulary balance (BALANCE)",
       title = "1) Simulated data") +
  guides(color = guide_legend(reverse = T, nrow = 1),
         linetype = guide_legend(reverse = T, nrow = 1)) 
```

```{r simulation1-plus_observed_data, echo = FALSE, cache = TRUE}
Observed_plot_inputs_TE <- keepers_ws_TE %>% 
  select(baby_id, ID_testdate, age_days, number_of_te, word_vocab, total_words_dom, total_words_nondom, balance_vocab) %>%
  pivot_longer(c(word_vocab, total_words_dom, total_words_nondom), names_to = "vocab_type", values_to = "number") %>%
  mutate(vocab_type = factor(vocab_type, levels=c("total_words_dom", "total_words_nondom", "word_vocab"))) %>%
  mutate(vocab_type = recode(vocab_type, 
                             total_words_dom = "Panel 2A: \n Dominant vocabulary \n (DOM)", 
                             total_words_nondom = "Panel 2B:\n Non-dominant vocabulary \n (NONDOM)",
                             word_vocab = "Panel 2C: \n Total vocabulary \n (WORD)")) %>%
  mutate(bin_BALANCE = cut(balance_vocab, breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5), include.lowest=T,
                     labels = c("0.1", "0.2", "0.3", "0.4", "0.5"))) %>%
  ggplot(aes(x = number, y = number_of_te, color = bin_BALANCE, linetype = bin_BALANCE)) +
  #geom_point() +
  stat_smooth(method = lm, se = F) +
  scale_color_manual(values=c("#bdbdbd", "#737373", "#525252", "#252525", "#000000")) + 
  scale_linetype_manual(values=c("longdash", "dotdash", "dashed", "twodash", "solid")) + 
  theme_minimal() + 
  facet_grid(. ~ vocab_type) +
  theme(axis.text.x = element_text(angle = 90),
        plot.title = element_text(face = 'bold', size = 12, hjust = 0, vjust = 2),
        plot.title.position = "plot",
        text = element_text(size=12),
        legend.position = "bottom",
        panel.spacing.x = unit(2, "lines"),
        panel.background = element_rect(fill = NA, color = "black")) +
  ylim(0, 600) +
  xlim(0, 1000) +
  labs(x = "Number of words", 
       y = "Number of translation equivalents (TE)", 
       color = "Vocabulary balance (BALANCE)",
       linetype = "Vocabulary balance (BALANCE)",
       title = "2) Observed data") +
  guides(colour = guide_legend(reverse = T, nrow = 1),
         linetype = guide_legend(reverse = T, nrow = 1))

@@

```{r fig1, fig.cap="Number of translation equivalents (TE) across different levels of vocabulary balance (BALANCE) in relation to dominant vocabulary size (DOM; Panel A), non-dominant vocabulary size (NONDOM; Panel B), and word vocabulary (WORD; Panel C). Row 1 represents the simulated data in Study 1 while holding the number of learnable words (LEARNABLE) constant at 600 and BIAS constant at 1. Row 2 represents the observed vocabulary data in Study 2.", echo=FALSE, dpi=600, fig.align='center', fig.height=7, out.height="80%", out.width="110%", fig.pos = "H", cache = TRUE}

@@

Translation equivalent knowledge was simulated across children at four developmental levels (the number of LEARNABLE words = 100, 200, 300, 400, 500, 600), in conjunction with a wide range of values for words in the dominant language (DOM) and the non-dominant language (NONDOM). In total, data from 301 simulated children were generated (see Table 2 for a summary of the parameters used in this simulation). Again, balance (BALANCE) was calculated based on the values of DOM and NONDOM. We also calculated the number of singlet words in the dominant (DOM-SINGLET) and non-dominant (NONDOM-SINGLET) languages by the formula DOM-SINGLET = DOM - TE and NONDOM-SINGLET = NONDOM - TE. Figure 2 plots the number of TE, DOM-SINGLET, and NONDOM-SINGLET at different BALANCE levels across different levels of LEARNABLE. Across the three Panels (1A, 1B, and 1C),  simulated children at a later developmental level had larger concept vocabularies (CONCEPT). In Panel 1A, we continued to observe a pattern reported in prediction 1a, whereby simulated children with more balanced vocabularies produced more translation equivalents (TE). Moreover, regardless of balance, simulated children at later developmental levels (i.e., older children with more potentially LEARNABLE words) acquired more translation equivalents (TE). Panels 1B and 1C show that simulated children at later developmental levels also produced more singlets in the dominant and non-dominant languages. Overall, we generated 4 additional predictions (Prediction Set 2) made by the Bilingual Vocabulary Model. Compared to children at an earlier developmental level (i.e., younger infants with fewer potentially learnable words), children at a later developmental level (i.e., older infants with more potentially learnable words) will

* Prediction 2a: Have larger concept vocabularies (CONCEPT). 
* Prediction 2b: Produce more translation equivalents (TE), regardless of vocabulary balance (BALANCE).
* Prediction 2c: Produce more dominant-language singlet words (DOM-SINGLET). Moreover, those with the least _balanced_ vocabulary (BALANCE) will produce the most DOM-SINGLET.
* Prediction 2d: Produce more non-dominant-language singlets (NONDOM-SINGLET). Moreover, those with the most _balanced_ vocabulary (BALANCE) will produce the most NONDOM-SINGLET.

@@

## visualizing Simulation 2
simulation2 <- data_simulation2 %>%
  pivot_longer(-c(LEARNABLE, WORD, CONCEPT, BALANCE, bin_BALANCE), names_to = "vocab_type", values_to = "number") %>%
  filter(vocab_type %notin% c("DOM", "NONDOM")) %>%
  mutate(vocab_type = factor(vocab_type, levels=c("TE", "DOM_SINGLET", "NONDOM_SINGLET")),
         vocab_type = recode(vocab_type, 
                             TE = "Panel 1A: \n Translation equivalents \n (TE)",
                             DOM_SINGLET = "Panel 1B: \n Singlets in dominant language \n (DOM_SINGLET)",
                             NONDOM_SINGLET = "Panel 1C: \n Singlets in non-dominant language \n (NONDOM_SINGLET)")) %>%
  ggplot(aes(x = LEARNABLE, y = number, color = bin_BALANCE, linetype = bin_BALANCE)) +
  facet_grid(. ~ vocab_type) +
  stat_smooth(method = lm, se = F) + 
  scale_color_manual(values=c("#bdbdbd", "#737373", "#525252", "#252525", "#000000")) + 
  scale_linetype_manual(values=c("longdash", "dotdash", "dashed", "twodash", "solid")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = 'bold', size = 12, hjust = 0, vjust = 2),
        plot.title.position = "plot",
        text = element_text(size=12),
        panel.spacing.x = unit(2, "lines"),
        panel.background = element_rect(fill = NA, color = "black")) + 
  guides(fill=guide_legend(nrow=3,byrow=TRUE)) + 
  labs(x = "\nNumber of learnable words (LEARNABLE)", 
       y = "Number of concepts (CONCEPT)", 
       color = "Vocabulary balance (BALANCE)",
       linetype = "Vocabulary balance (BALANCE)",
       title = "1) Simulated data") +
  guides(color = guide_legend(reverse = T, nrow = 1),
         linetype = guide_legend(reverse = T, nrow = 1)) 
```

```{r simulation2-plus_observed_data, echo = FALSE, cache = TRUE}
Observed_plot_LEARNABLE <- keepers_ws_TE %>%
  select(baby_id, age_months_percentile, Average_90percentile, balance_vocab, word_vocab, number_of_te, total_singlet_dom, total_singlet_nondom) %>%
  mutate(bin_BALANCE = cut(balance_vocab, breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5), include.lowest=T,
                     labels = c("0.1", "0.2", "0.3", "0.4", "0.5"))) %>%
  mutate(TE = number_of_te) %>%
  pivot_longer(c(total_singlet_nondom, total_singlet_dom, number_of_te), names_to = "vocab_type", values_to = "number_of_words") %>%
  mutate(vocab_type = factor(vocab_type, levels=c("number_of_te", "total_singlet_dom", "total_singlet_nondom")),
         vocab_type = recode(vocab_type, 
                             number_of_te = "Panel 2A: \n Translation equivalents \n (TE)",
                             total_singlet_dom = "Panel 2B: \n Singlets in dominant language \n (DOM_SINGLET)",
                             total_singlet_nondom = "Panel 2C: \n Singlets in non-dominant language \n (NONDOM_SINGLET)")) %>%
  ggplot(aes(x = age_months_percentile, y = number_of_words, color = bin_BALANCE, linetype = bin_BALANCE)) +
  facet_grid(. ~ vocab_type) +
  stat_smooth(method = lm, se = F) + 
  scale_color_manual(values=c("#bdbdbd", "#737373", "#525252", "#252525", "#000000")) + 
  scale_linetype_manual(values=c("longdash", "dotdash", "dashed", "twodash", "solid")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = 'bold', size = 12, hjust = 0, vjust = 2),
        plot.title.position = "plot",
        text = element_text(size=12),
        panel.spacing.x = unit(2, "lines"),
        panel.background = element_rect(fill = NA, color = "black")) + 
  guides(fill=guide_legend(nrow=3,byrow=TRUE)) + 
  ylim(0, 450) +
  #xlim(200, 700) +
  scale_x_continuous(breaks = seq(18, 30, by = 2)) +
  labs(x = "\nAge (in months)", 
       y = "Number of concepts (CONCEPT)", 
       color = "Vocabulary balance (BALANCE)",
       linetype = "Vocabulary balance (BALANCE)",
       title = "2) Observed data") +
  guides(color = guide_legend(reverse = T, nrow = 1),
         linetype = guide_legend(reverse = T, nrow = 1))

@@

```{r fig2, fig.cap="Number of translation equivalents (TE; Panel A), singlets in dominant (DOM-SINGLET; Panel B) and singlets in the non-dominant language (NONDOM-SINGLET; Panel C) across different levels of vocabulary balance (BALANCE) in relation to different developmental levels/ages, which sets the number of LEARNABLE words. Row 1 represents the simulated data in Study 1. Row 2 represents the observed vocabulary data in Study 2.", echo=FALSE, dpi=600, fig.align='center', fig.height=7, out.width="80%", fig.pos = "H", cache = TRUE}

@@

Prediction 1b was that children with larger word vocabularies and larger dominant-language vocabularies would produce more translation equivalents, and the results from our dataset confirmed this prediction, for word vocabulary (WORD): _r_(`r cor_TE_word$parameter[1]`) = `r round(cor_TE_word$estimate[1],2)`, $p `r papaja::printp(cor_TE_word$p.value[1])`$, and dominant-language vocabulary (DOM): _r_(`r cor_TE_dom$parameter[1]`) = `r round(cor_TE_dom$estimate[1],2)`, $p `r papaja::printp(cor_TE_dom$p.value[1])`$. Figure 1 Row 2 further illustrates these relationships observed in our dataset.

Prediction 1c was that children who produce more words in the non-dominant language (NONDOM) would produce more translation equivalents (TE), specifically that this relationship would be nearly perfect. As shown in Figure 1 Row 2 (Observed data), we observed that these two variables were indeed nearly perfectly correlated, _r_(`r cor_TE_nondom$parameter[1]`) = `r round(cor_TE_nondom$estimate[1],2)`, $p `r papaja::printp(cor_TE_nondom$p.value[1])`$.

@@

## 2.2.3 Testing Prediction Set 2: The vocabulary composition of bilingual children at different developmental levels
Prediction Set 2 pertained to expected patterns of acquisition of translation equivalents and singlets for children of different developmental levels. In our data set, developmental level was approximated by children’s age. Figure 2 Panel B shows the number of translation equivalents (TE), singlets in the dominant language (DOM-SINGLET) and in the non-dominant language (NONDOM-SINGLET) of the bilingual children as a function of different ages (a proxy for developmental level), used to estimate the number of LEARNABLE words.

@@

Prediction 2a was that older children (i.e., those at a later developmental level) would have larger concept vocabularies than younger children (i.e., those at an earlier developmental level). We observed a positive correlation between age (used as a proxy for developmental level, which determines LEARNABLE) and concept vocabulary (CONCEPT) in our dataset, _r_(`r cor_age_CONCEPT$parameter[1]`) = `r round(cor_age_CONCEPT$estimate[1],2)`, $p `r papaja::printp(cor_age_CONCEPT$p.value[1])`$, and therefore confirmed the prediction.

Prediction 2b was that older children would produce more translation equivalents than younger children. We observed a positive correlation between age (our proxy for LEARNABLE) and number of translation equivalents in our dataset, _r_(`r cor_age_TE$parameter[1]`) = `r round(cor_age_TE$estimate[1],2)`, $p `r papaja::printp(cor_age_TE$p.value[1])`$, and therefore confirmed the prediction.

Prediction 2c was that both older children and those with the least balanced vocabularies (BALANCE) would produce more dominant-language singlets (DOM-SINGLET). This pattern was confirmed by the results from our dataset, with a positive correlation between dominant-language singlets (DOM-SINGLET) and age (which determined LEARNABLE), _r_(`r cor_age_DOMSING$parameter[1]`) = `r round(cor_age_DOMSING$estimate[1],2)`, $p `r papaja::printp(cor_age_DOMSING$p.value[1])`$, and a negative correlation between BALANCE and dominant-language singlets (DOM-SINGLET), _r_(`r cor_BALANCE_DOMSING$parameter[1]`) = `r round(cor_BALANCE_DOMSING$estimate[1],2)`, $p `r papaja::printp(cor_BALANCE_DOMSING$p.value[1])`$. As illustrated in Figure 2 Panel 2B, children with the least balanced vocabulary of 0.1 produced the most singlets in their dominant language (mean = `r round(mean_DOMSINGLET_BALANCEsubset$mean_DOM_SINGLET[1],0)`), whereas those with the most balanced vocabulary of 0.5 produced the fewest (mean = `r round(mean_DOMSINGLET_BALANCEsubset$mean_DOM_SINGLET[5],0)`).

Prediction 2d was that older children and those with the most balanced vocabularies (BALANCE) would produce more singlets in their non-dominant language. This pattern was also observed in our dataset, with a positive correlation between the number of non-dominant singlets (NONDOM-SINGLET) and age (which determined LEARNABLE), _r_(`r cor_age_NONDOMSING$parameter[1]`) = `r round(cor_age_NONDOMSING$estimate[1],2)`, $p = `r papaja::printp(cor_age_NONDOMSING$p.value[1])`$, and a positive correlation between BALANCE and the number of non-dominant singlets (NONDOM-SINGLET), _r_(`r cor_BALANCE_NONDOMSING$parameter[1]`) = `r round(cor_BALANCE_NONDOMSING$estimate[1],2)`, $p `r papaja::printp(cor_BALANCE_NONDOMSING$p.value[1])`$. As shown in Figure 2 Panel 2C, we observed that children produced very few singlets in their non-dominant language, although the most balanced children with BALANCE of 0.5 produced the most singlets in their non-dominant language (mean = `r round(mean_NONDOMSINGLET_BALANCEsubset$mean_NONDOM_SINGLET[5],0)`) and the least balanced children with BALANCE of 0.1 produced the fewest (mean = `r round(mean_NONDOMSINGLET_BALANCEsubset$mean_NONDOM_SINGLET[1],0)`).

@@

>>>>

* Prediction 1a: Children with more balanced vocabularies (BALANCE) will produce more translation equivalents (TE) (Figure 1, Panel 1A).
* Prediction 1b: Children who produce more total words (WORD) or more dominant-language words (DOM) will produce more translation equivalents (TE) (Figure 1, Panels 1A and 1B). 
* Prediction 1c: Children who produce more non-dominant language words (NONDOM) will produce more translation equivalents (TE). Unlike for WORD and DOM this will not interact with BALANCE; instead, non-dominant vocabulary size will be an almost perfect predictor of translation equivalent knowledge (Figure 1, Panel 1C).

@@

## visualizing Simulation 1
simulation1 <- data_simulation1 %>%
  # convert to long data
  pivot_longer(c(WORD, DOM, NONDOM), names_to = "vocab_type", values_to = "number") %>%
  mutate(vocab_type = factor(vocab_type, levels=c("WORD", "DOM", "NONDOM"))) %>%
  mutate(vocab_type = recode(vocab_type, 
                             WORD = "Panel 1A: \n Total vocabulary \n (WORD)",
                             DOM = "Panel 1B: \n Dominant vocabulary \n (DOM)", 
                             NONDOM = "Panel 1C: \n Non-dominant vocabulary \n (NONDOM)")) %>%
  ggplot(aes(x = number, y = TE, color = bin_BALANCE, linetype = bin_BALANCE)) +
  stat_smooth(method = lm, se = F) +
  scale_color_manual(values=c("#bdbdbd", "#737373", "#525252", "#252525", "#000000")) + 
  scale_linetype_manual(values=c("longdash", "dotdash", "dashed", "twodash", "solid")) +
  theme_minimal() + 
  facet_grid(. ~ vocab_type) +
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom",
        plot.title = element_text(face = 'bold', size = 12, hjust = 0, vjust = 2),
        plot.title.position = "plot",
        text = element_text(size=12),
        panel.spacing.x = unit(2, "lines"),
        panel.background = element_rect(fill = NA, color = "black")) +
  ylim(0, 600) +
  xlim(0, 1000) +
  labs(x = "Number of words", 
       y = "Number of translation equivalents (TE)", 
       color = "Vocabulary balance (BALANCE)",
       linetype = "Vocabulary balance (BALANCE)",
       title = "1) Simulated data") +
  guides(color = guide_legend(reverse = T, nrow = 1),
         linetype = guide_legend(reverse = T, nrow = 1)) 
```

```{r simulation1-plus_observed_data, echo = FALSE, cache = TRUE}
Observed_plot_inputs_TE <- keepers_ws_TE %>% 
  select(baby_id, ID_testdate, age_days, number_of_te, word_vocab, total_words_dom, total_words_nondom, balance_vocab) %>%
  pivot_longer(c(word_vocab, total_words_dom, total_words_nondom), names_to = "vocab_type", values_to = "number") %>%
  mutate(vocab_type = factor(vocab_type, levels=c("word_vocab", "total_words_dom", "total_words_nondom"))) %>%
  mutate(vocab_type = recode(vocab_type, 
                             word_vocab = "Panel 2A: \n Total vocabulary \n (WORD)",
                             total_words_dom = "Panel 2B: \n Dominant vocabulary \n (DOM)", 
                             total_words_nondom = "Panel 2C:\n Non-dominant vocabulary \n (NONDOM)")) %>%
  mutate(bin_BALANCE = cut(balance_vocab, breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5), include.lowest=T,
                     labels = c("0.1", "0.2", "0.3", "0.4", "0.5"))) %>%
  ggplot(aes(x = number, y = number_of_te, color = bin_BALANCE, linetype = bin_BALANCE)) +
  #geom_point() +
  stat_smooth(method = lm, se = F) +
  scale_color_manual(values=c("#bdbdbd", "#737373", "#525252", "#252525", "#000000")) + 
  scale_linetype_manual(values=c("longdash", "dotdash", "dashed", "twodash", "solid")) + 
  theme_minimal() + 
  facet_grid(. ~ vocab_type) +
  theme(axis.text.x = element_text(angle = 90),
        plot.title = element_text(face = 'bold', size = 12, hjust = 0, vjust = 2),
        plot.title.position = "plot",
        text = element_text(size=12),
        legend.position = "bottom",
        panel.spacing.x = unit(2, "lines"),
        panel.background = element_rect(fill = NA, color = "black")) +
  ylim(0, 600) +
  xlim(0, 1000) +
  labs(x = "Number of words", 
       y = "Number of translation equivalents (TE)", 
       color = "Vocabulary balance (BALANCE)",
       linetype = "Vocabulary balance (BALANCE)",
       title = "2) Observed data") +
  guides(colour = guide_legend(reverse = T, nrow = 1),
         linetype = guide_legend(reverse = T, nrow = 1))

@@

```{r fig1, fig.cap="Number of translation equivalents (TE) across different levels of vocabulary balance (BALANCE) in relation to word vocabulary (WORD; Panel A), dominant vocabulary size (DOM; Panel B), and non-dominant vocabulary size (NONDOM; Panel C). Row 1 represents the simulated data in Study 1 while holding the number of learnable words (LEARNABLE) constant at 600 and BIAS constant at 1. Row 2 represents the observed vocabulary data in Study 2.", echo=FALSE, dpi=600, fig.align='center', fig.height=7, out.height="80%", out.width="110%", fig.pos = "H", cache = TRUE}

@@

Translation equivalent knowledge was simulated across children at four developmental levels (the number of LEARNABLE words = 100, 200, 300, 400, 500, 600), in conjunction with a wide range of values for words in the dominant language (DOM) and the non-dominant language (NONDOM). In total, data from 301 simulated children were generated (see Table 2 for a summary of the parameters used in this simulation). Again, balance (BALANCE) was calculated based on the values of DOM and NONDOM. We also calculated the number of singlet words in the dominant (DOM-SINGLET) and non-dominant (NONDOM-SINGLET) languages by the formula DOM-SINGLET = DOM - TE and NONDOM-SINGLET = NONDOM - TE. Figure 2 plots the number of CONCEPT, TE, DOM-SINGLET, and NONDOM-SINGLET at different BALANCE levels across different levels of LEARNABLE. As shown in Panel 1A, simulated children at a later developmental level had larger concept vocabularies (CONCEPT). In Panel 1B, we continued to observe a pattern reported in prediction 1a, whereby simulated children with more balanced vocabularies produced more translation equivalents (TE). Moreover, regardless of balance, simulated children at later developmental levels (i.e., older children with more potentially LEARNABLE words) acquired more translation equivalents (TE). Panels 1C and 1D show that simulated children at later developmental levels also produced more singlets in the dominant and non-dominant languages. Overall, we generated 4 additional predictions (Prediction Set 2) made by the Bilingual Vocabulary Model. Compared to children at an earlier developmental level (i.e., younger infants with fewer potentially learnable words), children at a later developmental level (i.e., older infants with more potentially learnable words) will

* Prediction 2a: Have larger concept vocabularies (CONCEPT). (Figure 2, Panel 1A)
* Prediction 2b: Produce more translation equivalents (TE), regardless of vocabulary balance (BALANCE). (Figure 2, Panel 1B)
* Prediction 2c: Produce more dominant-language singlet words (DOM-SINGLET). Moreover, those with the least _balanced_ vocabulary (BALANCE) will produce the most DOM-SINGLET. (Figure 2, Panel 1C)
* Prediction 2d: Produce more non-dominant-language singlets (NONDOM-SINGLET). Moreover, those with the most _balanced_ vocabulary (BALANCE) will produce the most NONDOM-SINGLET. (Figure 2, Panel 1D)

@@

## visualizing Simulation 2
simulation2 <- data_simulation2 %>%
  pivot_longer(-c(LEARNABLE, WORD, BALANCE, bin_BALANCE), names_to = "vocab_type", values_to = "number") %>%
  filter(vocab_type %notin% c("DOM", "NONDOM")) %>%
  mutate(vocab_type = factor(vocab_type, levels=c("CONCEPT", "TE", "DOM_SINGLET", "NONDOM_SINGLET")),
         vocab_type = recode(vocab_type, 
                             CONCEPT = "Panel 1A: \n Concept vocabulary \n (CONCEPT)",
                             TE = "Panel 1B: \n Translation equivalents \n (TE)",
                             DOM_SINGLET = "Panel 1C: \n Singlets in dominant language \n (DOM_SINGLET)",
                             NONDOM_SINGLET = "Panel 1D: \n Singlets in non-dominant language \n (NONDOM_SINGLET)")) %>%
  ggplot(aes(x = LEARNABLE, y = number, color = bin_BALANCE, linetype = bin_BALANCE)) +
  facet_grid(. ~ vocab_type) +
  stat_smooth(method = lm, se = F) + 
  scale_color_manual(values=c("#bdbdbd", "#737373", "#525252", "#252525", "#000000")) + 
  scale_linetype_manual(values=c("longdash", "dotdash", "dashed", "twodash", "solid")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = 'bold', size = 10, hjust = 0, vjust = 2),
        plot.title.position = "plot",
        text = element_text(size=10),
        panel.spacing.x = unit(2, "lines"),
        panel.background = element_rect(fill = NA, color = "black")) + 
  guides(fill=guide_legend(nrow=3,byrow=TRUE)) + 
  labs(x = "\nNumber of learnable words (LEARNABLE)", 
       y = "Number of concepts (CONCEPT)", 
       color = "Vocabulary balance (BALANCE)",
       linetype = "Vocabulary balance (BALANCE)",
       title = "1) Simulated data") +
  guides(color = guide_legend(reverse = T, nrow = 1),
         linetype = guide_legend(reverse = T, nrow = 1)) 
```

```{r simulation2-plus_observed_data, echo = FALSE, cache = TRUE}
Observed_plot_LEARNABLE <- keepers_ws_TE %>%
  select(baby_id, age_months_percentile, Average_90percentile, balance_vocab, 
         word_vocab, concept_vocab, number_of_te, total_singlet_dom, total_singlet_nondom) %>%
  mutate(bin_BALANCE = cut(balance_vocab, breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5), include.lowest=T,
                     labels = c("0.1", "0.2", "0.3", "0.4", "0.5"))) %>%
  pivot_longer(c(total_singlet_nondom, total_singlet_dom, number_of_te, concept_vocab), names_to = "vocab_type", values_to = "number_of_words") %>%
  mutate(vocab_type = factor(vocab_type, levels=c("concept_vocab", "number_of_te", "total_singlet_dom", "total_singlet_nondom")),
         vocab_type = recode(vocab_type, 
                             concept_vocab = "Panel 2A: \n Concept vocabulary \n (CONCEPT)",
                             number_of_te = "Panel 2B: \n Translation equivalents \n (TE)",
                             total_singlet_dom = "Panel 2C: \n Singlets in dominant language \n (DOM_SINGLET)",
                             total_singlet_nondom = "Panel 2D: \n Singlets in non-dominant language \n (NONDOM_SINGLET)")) %>%
  ggplot(aes(x = age_months_percentile, y = number_of_words, color = bin_BALANCE, linetype = bin_BALANCE)) +
  facet_grid(. ~ vocab_type) +
  stat_smooth(method = lm, se = F) + 
  scale_color_manual(values=c("#bdbdbd", "#737373", "#525252", "#252525", "#000000")) + 
  scale_linetype_manual(values=c("longdash", "dotdash", "dashed", "twodash", "solid")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = 'bold', size = 10, hjust = 0, vjust = 2),
        plot.title.position = "plot",
        text = element_text(size=10),
        panel.spacing.x = unit(2, "lines"),
        panel.background = element_rect(fill = NA, color = "black")) + 
  guides(fill=guide_legend(nrow=3,byrow=TRUE)) + 
  ylim(0, 450) +
  scale_x_continuous(breaks = seq(18, 30, by = 2)) +
  labs(x = "\nAge (in months)", 
       y = "Number of concepts (CONCEPT)", 
       color = "Vocabulary balance (BALANCE)",
       linetype = "Vocabulary balance (BALANCE)",
       title = "2) Observed data") +
  guides(color = guide_legend(reverse = T, nrow = 1),
         linetype = guide_legend(reverse = T, nrow = 1))

@@

```{r fig2, fig.cap="Number of concept vocabulary (CONCEPT; Panel A), translation equivalents (TE; Panel B), singlets in dominant (DOM-SINGLET; Panel C) and singlets in the non-dominant language (NONDOM-SINGLET; Panel D) across different levels of vocabulary balance (BALANCE) in relation to different developmental levels/ages, which sets the number of LEARNABLE words. Row 1 represents the simulated data in Study 1. Row 2 represents the observed vocabulary data in Study 2.", echo=FALSE, dpi=600, fig.align='center', fig.height=7, out.width="100%", fig.pos = "H", cache = TRUE}

@@

Prediction 1b was that children with larger word vocabularies and larger dominant-language vocabularies would produce more translation equivalents, and the results from our dataset confirmed this prediction, for word vocabulary (WORD): _r_(`r cor_TE_word$parameter[1]`) = `r round(cor_TE_word$estimate[1],2)`, $p `r papaja::printp(cor_TE_word$p.value[1])`$, and dominant-language vocabulary (DOM): _r_(`r cor_TE_dom$parameter[1]`) = `r round(cor_TE_dom$estimate[1],2)`, $p `r papaja::printp(cor_TE_dom$p.value[1])`$. Figure 1 Panels 2A and 2B further illustrates these relationships observed in our dataset.

Prediction 1c was that children who produce more words in the non-dominant language (NONDOM) would produce more translation equivalents (TE), specifically that this relationship would be nearly perfect. As shown in Figure 1 Panel 2C, we observed that these two variables were indeed nearly perfectly correlated, _r_(`r cor_TE_nondom$parameter[1]`) = `r round(cor_TE_nondom$estimate[1],2)`, $p `r papaja::printp(cor_TE_nondom$p.value[1])`$.

@@

## 2.2.3 Testing Prediction Set 2: The vocabulary composition of bilingual children at different developmental levels
Prediction Set 2 pertained to expected patterns of acquisition of translation equivalents and singlets for children of different developmental levels. In our data set, developmental level was approximated by children’s age. Figure 2 Panel B shows the number of concept vocabulary (CONCEPT), translation equivalents (TE), singlets in the dominant language (DOM-SINGLET) and in the non-dominant language (NONDOM-SINGLET) of the bilingual children as a function of different ages (a proxy for developmental level), used to estimate the number of LEARNABLE words.

@@

Prediction 2a was that older children (i.e., those at a later developmental level) would have larger concept vocabularies than younger children (i.e., those at an earlier developmental level). As shown in Figure 2 Panel 2A, we observed a positive correlation between age (used as a proxy for developmental level, which determines LEARNABLE) and concept vocabulary (CONCEPT) in our dataset, _r_(`r cor_age_CONCEPT$parameter[1]`) = `r round(cor_age_CONCEPT$estimate[1],2)`, $p `r papaja::printp(cor_age_CONCEPT$p.value[1])`$, and therefore confirmed the prediction.

Prediction 2b was that older children would produce more translation equivalents than younger children. As illustrated in Figure 2 Panel 2B, we observed a positive correlation between age (our proxy for LEARNABLE) and number of translation equivalents in our dataset, _r_(`r cor_age_TE$parameter[1]`) = `r round(cor_age_TE$estimate[1],2)`, $p `r papaja::printp(cor_age_TE$p.value[1])`$, and therefore confirmed the prediction.

Prediction 2c was that both older children and those with the least balanced vocabularies (BALANCE) would produce more dominant-language singlets (DOM-SINGLET). This pattern was confirmed by the results from our dataset, with a positive correlation between dominant-language singlets (DOM-SINGLET) and age (which determined LEARNABLE), _r_(`r cor_age_DOMSING$parameter[1]`) = `r round(cor_age_DOMSING$estimate[1],2)`, $p `r papaja::printp(cor_age_DOMSING$p.value[1])`$, and a negative correlation between BALANCE and dominant-language singlets (DOM-SINGLET), _r_(`r cor_BALANCE_DOMSING$parameter[1]`) = `r round(cor_BALANCE_DOMSING$estimate[1],2)`, $p `r papaja::printp(cor_BALANCE_DOMSING$p.value[1])`$. As illustrated in Figure 2 Panel 2C, children with the least balanced vocabulary of 0.1 produced the most singlets in their dominant language (mean = `r round(mean_DOMSINGLET_BALANCEsubset$mean_DOM_SINGLET[1],0)`), whereas those with the most balanced vocabulary of 0.5 produced the fewest (mean = `r round(mean_DOMSINGLET_BALANCEsubset$mean_DOM_SINGLET[5],0)`).

Prediction 2d was that older children and those with the most balanced vocabularies (BALANCE) would produce more singlets in their non-dominant language. This pattern was also observed in our dataset, with a positive correlation between the number of non-dominant singlets (NONDOM-SINGLET) and age (which determined LEARNABLE), _r_(`r cor_age_NONDOMSING$parameter[1]`) = `r round(cor_age_NONDOMSING$estimate[1],2)`, $p = `r papaja::printp(cor_age_NONDOMSING$p.value[1])`$, and a positive correlation between BALANCE and the number of non-dominant singlets (NONDOM-SINGLET), _r_(`r cor_BALANCE_NONDOMSING$parameter[1]`) = `r round(cor_BALANCE_NONDOMSING$estimate[1],2)`, $p `r papaja::printp(cor_BALANCE_NONDOMSING$p.value[1])`$. As shown in Figure 2 Panel 2D, we observed that children produced very few singlets in their non-dominant language, although the most balanced children with BALANCE of 0.5 produced the most singlets in their non-dominant language (mean = `r round(mean_NONDOMSINGLET_BALANCEsubset$mean_NONDOM_SINGLET[5],0)`) and the least balanced children with BALANCE of 0.1 produced the fewest (mean = `r round(mean_NONDOMSINGLET_BALANCEsubset$mean_NONDOM_SINGLET[1],0)`).

@@

Finally, the bilingual children in our sample were learning one particular language pair: French and English. In both French and English, children tend to initially learn more nouns than verbs or other word classes (i.e., noun bias; Gentner, 1982), whereas children learning other languages such as Mandarin show more equal learning of nouns and verbs (Tardif, 1996). Moreover, French and English are also typologically similar, for example they share the same basic subject-verb-object word order. Similarities across bilinguals’ two languages might make it more likely that children will acquire translation equivalents. Future research could extend our approach to bilinguals learning different language pairs, and could also examine translation equivalent learning separately for different word classes.
